A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-05 doc/large_strain.tex
Improve some wordings and fox typos

\usepackage{pgfkeys,pgfmath,pgfcore}
% \pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfkeys{/pgf/number format/fixed,/pgf/number format/precision=3}

@@

\newcommand{\changeJP}[1]{\textcolor{blue}{#1}}
\newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}

@@

    \changeJP{To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
    In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
    Specifically, we study different numerical implementations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
    }
    In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
    \changeJP{We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
  \end{abstract}

@@

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

In this work, we perform \changeJP{a preliminary investigation of} different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
and performance {\color{red} on a single node of a high performance computing cluster.}
\changeJP{
The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a \changeJP{generalized matrix-free framework that incorporates SIMD vectorization} and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
}
In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
and employ them to define a \mbox{geometric} multigrid preconditioner.

The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics. 
\changeJP{
Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
In \ref{sec:mf} we describe the different matrix-free operator evaluation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for finite-strain hyperelasticity.
}
The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}. 
\changeJP{Lastly, the} results are summarized in Section \ref{sec:summary}.

@@

\subsection{Constitutive modelling
\label{sec: Constitutive modelling}
}

For the current study, we use the compressible Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively.
\changeJP{
Derived using statistical thermodynamics applied to cross-linked polymers, the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
}
It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated fourth-order material part of the spatial tangent tensor are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor
with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

@@

\section{\changeJP{Description of numerical framework}
\label{sec: Description of numerical framework}
}

The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
This library offers a number of paradigms by which to parallelize a finite-element code.
In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fine_level_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_fine_level}
\end{figure}
%
%
Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.

In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged; 
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

{\color{red}
Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
\changeJP{They are introduced in order of lowest memory footprint to largest:}
}

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization \;
      distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache scalar quantities only}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache material part of the fourth-order spatial tangent stiffness tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

@@

\todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}

>>>>

@@

@@

    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic finite element tangent operator numerically.
    In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
    and employ them to define a geometric multigrid preconditioner.
    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
  \end{abstract}

@@

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
{\color{red} which is also limited to small strain linear elasticity.}\todo{omit?}
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference?}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
and performance {\color{red} on a single node of an higher performance computing cluster.}
In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
and employ them to define a \mbox{geometric} multigrid preconditioner.
MPI parallelization together with SIMD vectorization
{\color{red}
are provided by the \texttt{deal.II} \cite{dealII90} finite element library, used for this study.
}

The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics. Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches. In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain hyperelasticity. The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}. The results are summarized in Section \ref{sec:summary}.

@@

\subsection{Constitutive modelling}

For the current study, we use the compressible Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated fourth-order material part of the spatial tangent tensor are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor
with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

@@

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

{\color{red}
Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
}

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization \;
      distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache scalar quantities only}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        {\color{red}extract local vector values on this element $\gz x_K$ } \;
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache material part of the fourth-order spatial tangent stiffness tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

@@
